---
title: "KPPA CART"
author: "Fabian Bong"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Abstract

Kurtosis Projection Pursuit with Classification and Regression Trees (KPPA Cart) encompasses a framework for the robust, unsupervised classification of integrated Multi-Omics data based on kurtosis as the function of interest. This introductory vignette provides an overview of the steps of the workflow, installation, and source of the data set used as example.

## Contents

Installation

Introduction to KPPACart

More Info & Citation

Example

## Installation

The latest version of the library can be installed from github using `devtools`.

```{r}
install.packages("devtools")
devtools::install_github("FabianBong/KPPACart")
```

## Introduction to KPPACart

KPPACart has 2 basic goals: find unsupervised clusters in the data and display the features that cause the clustering. The algorithm achieves the goals by randomly sampling `n_features` from the data `X`. Then kurtosis projection pursuit is applied to the subset of data. The resulting clusters are given a label based on the k-means algorithm and the variable `exp_clusters`. A random forest is used to predict the assigned labels based on the subset data; thus providing importance scores for all features in the subset. This procedure is repeated `n_iterations` times. When finished, the importance scores are averaged for each feature and a last kurtosis projection pursuit is run on the top `n_features` features, leading to the final result.

## More Info & Citation

Here goes the citation for the paper.

## Example

A sample data set can be loaded by using `data("KPPACart.Data")`. It contains a list with X, Age and Sex. X represents an RNA count matrix, Age and Sex contain numerically coded age/sex groups. Since Age and Sex contain two groups each, the samples can be split into four total groups. To run KPPACart simply call the function with the default parameters or determine the parameters yourself. The power of KPPACart is that it shows better separation than application of PCA.

```{r}
# Load library
library("KPPACart")

# Load data
data("KPPACart.Data")

# Apply PCA
pca <- prcomp(t(KPPACart.Data$X))

# plot pca colored by age and sex
plot(pca$x[,1], pca$x[,2], col=factor(paste(KPPACart.Data$Sex,KPPACart.Data$Age)),
     xlab="PC1", ylab="PC2")

# Apply KPPACart
n_features <- 100
n_iterations <- 1000
k_dim <- 10
exp_clusters <- 4
n_cores <- 8
res <- KPPACart::KPPACart(KPPACart.Data$X,
                          n_features = n_features,
                          n_iterations = n_iterations,
                          k_dim = k_dim,
                          exp_clusters = exp_clusters,
                          n_cores = n_cores)

# Plot KPPACart result
plot(res$T[,1], res$T[,2], col=factor(paste(KPPACart.Data$Sex,KPPACart.Data$Age)),
     xlab="Component 1", ylab="Component 2")
```
